Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-01-22 ms.tex
revise TNBC again

In some of the analysis, we consider breast cancer. 
\citet{salehi2021clonal} studied multiple TNBC PDX clones (triple-negative breast cancer patient-derived xenografts).
Their results suggest that aneuploidy and other copy number alterations provid resistance to the drug cisplatin.  
Thus, we assume that aneuploid breast cancer cells are are partially or even completely resistant, i.e., $\Delta_m \ge \Delta_a > 0$, and that the division and death rates of aneuploid cells in the presence of cisplatin is the same as those of  sensitive breast cancer cells in the absence of the drug. 
The growth rate of breast cancer cells in the absence of drug is $0.0085$ per day \citep{spratt1996rates}.
The doubling time of breast cancer cells in the absence of cellular death is 8.2 days \citep{rew2000cell}.
Thus, the division rate of aneuploid cells is $\lambda_a=\log{(2)}/8.2=0.0845$ per day, and their death rate is $\mu_a=0.076$ per day.
To determine the division and death rates of sensitive cells, we chose TNBC-SA535 clone H from \citet{salehi2021clonal}, which has a relative fitness of $\approx 1.011$. Thus, the selection coefficient is the ratio of the aneuploid and sensitive Wrightian fitness values minus 1 \citep{chevin2011measuring}. The Malthusian fitness is the log of the Wrightian fitness \citep{Wu2013a}. Thus, $\Delta_s = \log{0.9975} -0.0025$. 
We assume cisplatin affects cancer cells by increasing their death rate but does not affect the their division rate, hence $\lambda_s=0.08$ and $\mu_s=0.0825$ per day.
A similar analysis for TNBC-SA609 clone A and TNBC-SA1035 clone H from \citet{salehi2021clonal} gives $\lambda_s=0.08, \mu_s=0.12$ and $\lambda_s=0.08, \mu=0.1$, respectively. % TODO it does not make sense that when you change aneuploid clones what changes is the sensitive growth rate - the sensitive growth rate should be constant and the aneuploid clone growth rate should change...

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $\lambda_s$ & Sensitive division rate& 0.08 & 1/days  & \citet{salehi2021clonal} \\ \hline
    $\mu_s$ & Sensitive death rate& $0.0825, 0.1, 0.12$ & 1/days  & \citet{salehi2021clonal} \\ \hline
    $\lambda_a$  & Aneuploid division rate & $0.0845$ & 1/days  & \citet{spratt1996rates,rew2000cell}  \\ \hline
    $\mu_a$ & Aneuploid death rate & $0.076$ & 1/days  & \citet{spratt1996rates,rew2000cell}  \\ \hline
  \end{tabular}
\caption{\textbf{Model parameters: Breast cancer.} 
Parameters not listed here remain unchanged from the values provided in \cref{table1}.
}
  \label{table2}
\end{center}
\end{table}

>>>>

We also consider breast cancer cells, relying on \citet{salehi2021clonal}, who studied multiple TNBC PDX clones (triple-negative breast cancer patient-derived xenografts.)
Their results suggest that aneuploidy and other copy number alterations provide resistance to the drug cisplatin.  
Thus, we assume that aneuploid breast cancer cells are partially or even fully drug-resistant, i.e., $\Delta_m \ge \Delta_a > 0$.
We chose three clones for which \cite{salehi2021clonal} report their relative Wrightian fitness, $1+s$, in the presence of the drug cisplatin: TNBC-SA609 clone A with $1+s=1.047$, TNBC-SA1035 clone H with $1+s=1.02$, and TNBC-SA535 clone H with $1+s=1.01$.
The growth rate $\Delta$, or the Malthusian fitness, is the log of the Wrightian fitness \citep{Wu2013a}. 
We therefore have the following set of equations: $0.047 = \exp{(\Delta_{SA609}-\Delta_s)}, 0.02=\exp{(\Delta_{SA1035}-\Delta_s)}, 0.01=\exp{(\Delta_{SA535}-\Delta_s)}$.
The growth rate of breast cancer cells in the absence of the drug is $0.0085$ per day \citep{spratt1996rates}.
The doubling time of breast cancer cells in the absence of cellular death is 8.2 days \citep{rew2000cell}.
Thus, the division rate of breast cancer cells in the absence of the drug is $\log{(2)}/8.2=0.0845$ per day, and their death rate is $0.076$ per day.
As all three clones grow in the presence of the drug, we set the growth rate of the least fit, TNBC-SA535 clone H, to the growth rate of breast cancer cells in the absence of the drug, $\Delta_{SA535}=0.0085$, and then solve the set of equations. % TODO needs to be the most fit, wait for Remus
We further assume the drug affects death rather than division. We therefore have $\lambda_s=\lambda_a=0.0845$ for all clones,  $\mu_{SA609} = 0.0415$, $\mu_{SA1035}=0.0667$, $\mu_{SA535}=0.085$, and $\mu_s=0.085$.

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units \\ \hline
    $\lambda_s$ & Sensitive division rate& 0.0848& 1/days \\ \hline
    $\mu_s$ & Sensitive death rate& $0.085$ & 1/days   \\ \hline
    $\lambda_a$  & Aneuploid division rate & $0.0845$ & 1/days    \\ \hline
    $\mu_a$ & Aneuploid death rate & $0.0415$, $0.0667$, $0.085$ & 1/days   \\ \hline
  \end{tabular}
\caption{\textbf{Model parameters: Breast cancer.} 
The derivation of these values, described in the methods section, uses fitness estimates from \citet{salehi2021clonal}.
Parameters not listed here remain unchanged from the values provided in \cref{table1}.
}
  \label{table2}
\end{center}
\end{table}
